---
title: "Constructing Prices - Category"
format: html
editor: visual
---

## Packages

For this first document, the following packages will be used:

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(magrittr)
library(gridExtra)

```

The `tidyverse` is a collection of R packages to read, visualize and manipulate data. `magrittr` is loaded for the `%<>%` operator. The package `gridExtra` allows us to easily exhibit multiple ggplots.

## Loading the data

```{r}
#| warning: false
#| message: false


prices<-read_csv("sell_prices.csv")
head(prices)

w<-read_csv("weights_validation.csv") 
head(w) 

```

Combinations of `item_id`, `dept_id`, `cat_id`, `store_id` and `state_id` provide the multiple hierarchical levels of interest of the forecasting competitions.

Attention: this code is for CATEGORY aggregate level. It needs to be adapted to other levels.

## Prices

We begin by aggregating the product/week prices across all stores as in the last script:

```{r}
#| warning: false 
#| message: false  

prices$cat_id<-unlist(str_split(prices$item_id, "_", 2, simplify = T), recursive = F)[,1]

w %<>% filter(Level_id == "Level12") %>%
  select(item_id = Agg_Level_1, store_id = Agg_Level_2, dollar_sales = Dollar_Sales)

prices %<>% left_join(w, by = c("item_id", "store_id"))
prices %<>% group_by(cat_id, item_id, wm_yr_wk) %>%
  summarise(sell_price = weighted.mean(sell_price, dollar_sales),
            dollar_sales = sum(dollar_sales)) %>%
  ungroup()
prices$sell_price[prices$dollar_sales == 0]<-0

head(prices)
```

```{r}
#| warning: false 
#| message: false 
#| echo: false  

rm(w) 
```

Below we create the weighted mean of weekly prices for each `cat_id`. Choose cat

```{r}
#| warning: false 
#| message: false  

# Summarizing 
prices %<>% group_by(cat_id, wm_yr_wk) %>%   
  summarise(price_wmean = weighted.mean(sell_price, dollar_sales),   
            price_median = median(sell_price)) %>%   
ungroup()  

head(prices) 
```
